<center> <h2>Exploratory Data Analysis - Week 1</h2> </center>
<center> <h3>---------------------------------------</h3> </center>

####*Video 1.1 Introduction*

**A. What the Class is About **

- this is about exploratory (pre-inference) analysis

####*Video 1.2 Principles of ANalytic Graphics*

**A. General Rules**

1. Show comparisons
    - evidence for a hypothesis is always relative to another, competing hypothesis
2. Show causality, mechanism, explanation, systematic structure
    - how is the system operating? How does the world work?
    - what is your "causal framework"?
3. Show multivariate data
    - show as much on a single plot as you can
    - "the real world is multivariate"
    - need to "escape flatland"
4. Integration of evidence
    - data graphics should make use of many modes of data presentation
    - don't let the tool (graphic device) drive the analysis
5. Describe and document the evidence with appropriate labels, scales, sources, etc.
    - a data graphic shoud tell a cmplete story that is credible
    - keep the code that made the plot
6. Content is king
    - if your story isn't interesting, graphs won't magically make it more interesting
    - start with the data and the story, then choose the appropriate plots
    
####*Video 1.3 Exploratory Graphs (pt 1)*

**A. What are Exploratory Graphs?**

- pretty much made for yourself
- why use them?
    - find patterns, understand properties, suggest modeling strategies
    - debug analysis
- made quickly, a large number are made
- the goal here is just for personal understanding of the data
- axes/legends cleaned up later
- color/size are primarily used for information

**B. Some Basic Tools**

- Five-number summary --> ```summary(x)```
- boxplots   --> ```boxplot(x, col="blue")```
- histograms --> 

```{r HistExample, echo=TRUE, eval=FALSE}
## Make a Histogram
hist(x, col="green", breaks=100)  ## breaks = "number of bars"
rug(x) ## plot the observations underneath the histogram

abline(v = 12, lwd = 2)  ## add a vertical line at 12, of width 2, to the histogram
abline(v = median(x), col = "magenta", lwd=4)  ## add another vertical line at the median, make it magenta, line width 4
```

- boxplot

```{r BoxPlotExample, echo=TRUE, eval=FALSE}
## Make a Boxplot with a horizontal line at the point of interest
boxplot(x, col="blue")
abline(h = 12) ## add a horizontal line at 12
```

- barplot --> 

```{r BarPlotExample, echo=TRUE, eval=FALSE}
## Simple Bar Plot
barplot(table(x), col = "wheat", main = "Number of Counties in Each Region")
```

####*Video 1.4 Exploratory Graphs (pt 2)*

**A. Simple Summaries of Data**

- Two dimensions
    - multiple/overlayed 1-D plots (Lattice/ggplot2)
    - scatterplots
    - smooth scatterplots
- >2 dimensions
    - overlayed/multiple 2-D plots; coplots
    - use color, size, shape to add dimensions
    - spinning plots
    - actual 3-D plots (not that useful)
    
**B. Some Examples**

- multiple boxplots

```{r MultiBoxExample, echo=TRUE, eval=FALSE}
## make side-by-side boxplots
boxplot(x ~ some_factor, data=dataset, col="red")
```

- stacked histograms

```{r MultiHistExample, echo=TRUE, eval=FALSE}
## make two histograms and stack them on each other
par(mfrow = c(2,1), mar = c(4,4,2,1)) ## change this to a two-row, 1 column plotting panel
hist(subset(x, region == "east"), col = "green"))
hist(subset(x, region == "west"), col = "green"))
```

- scatterplot with a horizontal dashed line

```{r DashLineExample, echo=TRUE, eval=FALSE}
## the scatterplot
with(pollution, plot(latitude, pm25, col = region)) ## you can color the points using some factor

## the dashed line
abline(h = 12, lwd = 2, lty = 2)
```

- panel of scatterplots

```{r BarPlotExample, echo=TRUE, eval=FALSE}
## set the graphical options
par(mfrow = c(1,2), mar = c(5,4,2,1))

## make each plot
with(subset(pollution, region == "west"), plot(latitude, pm25, main = "West"))
with(subset(pollution, region == "east"), plot(latitude, pm25, main = "east"))
```

**C. Summary**

- these plots are meant to be quick and dirty
- they let you summarize the data and highlight any broad features of interest
- you can explore basic questions and hypotheses (maybe ruling them out)
- see also [R Graph Gallery](http://rgraphgallery.blogspot.com/)
